Note on Linear Biases in MORUS

### 2018-05-01 morusAC.tex
Conference version. Specification was added and merged with a part of Notations.

\usepackage[hmargin=3cm,vmargin=2.8cm]{geometry}

@@

\section{Notation}

@@

Finally, let $x$ denote a register. Recall that registers of \cipher{MORUS} are composed of four words. We now define register-wise and word-wise rotations (denoted by \texttt{<<<} and \texttt{Rotl\_xxx\_yy} respectively in the \cipher{MORUS} design document):\\
\centerline{
\def\arraystretch{1.25}
\begin{tabular}{lcl}
$x \rotl y$ &:& circular shift of $x$ by $y$ bits to the left.\\
$x \rotlxy y$ &:& circular shift of each word within $x$ by $y$ bits to the left.\\
\end{tabular}
}

The $b_i$'s denote rotation constants, whose value depends on which version of \cipher{MORUS} is considered. Throughout, $w$ denotes the  word size, i.e. $w=32$ for \cipher{MORUS640} and $w=64$ for \cipher{MORUS1280}.

@@

>>>>

%\usepackage[hmargin=3cm,vmargin=2.8cm]{geometry}

@@

\section{Preliminaries}
\subsection{Specification of MORUS}
MORUS is a family of authenticated ciphers with two internal state sizes: 640 and 1280 bits. Each member is called \cipher{MORUS640} and \cipher{MORUS1280}, respectively. In the recommended parameters, \cipher{MORUS640} supports 128-bit keys and \cipher{MORUS1280} supports both 128-bit and 256-bit keys. The tag size is 128 bits or shorter, though the designers strongly recommend using a 128-bit tag.

In MORUS, the internal state is divided in two five registers. Let $q$ be the register size, where $q = 128$ for \cipher{MORUS640} and $q=256$ for \cipher{MORUS1280}. Let $S_i, i \in\{0,1,2,3,4\}$ be a register. Then, the internal state is represented as $S_0\|S_1\|S_2\|S_3\|S_4$. In MORUS, the internal state is first initialized by a key and nonce, and then updated by associated data blocks and message blocks step by step. Hence, we use a superscript to denote the step number of the state as $S^t_0\|S^t_1\|S^t_2\|S^t_3\|S^t_4$, where $t$ is 0 before the initialization is performed.

The computation of MORUS consists of the four parts; initialization, associated data process, encryption, and finalization. All parts share the core design of MORUS called the {\tt StepUpdate} function. Hereafter we explain {\tt StepUpdate}.

\subsubsection{Step Update Function.}
{\tt StepUpdate} takes as input the internal state $S^t = S^t_0\|S^t_1\|S^t_2\|S^t_3\|S^t_4$ and additional register-wise value $m^t$. $m^t$ is supposed to be associated data block or message block. $m^t$ can sometimes be 0 or constant when the additional input is unnecessary.

{\tt StepUpdate} is composed of 5 rounds with similar operations to update
the state $S^t$, in which the additional input $m^t$ is used in rounds 2 to 5, but not in round 1. Each round use the word-wise left circular shift, which divides a $q$-bit register value into 4 words of $q/4$ bits, and performs left circular shift operation for every $q/4$-bit word. Throughout the paper, we denote the word size by $w$, i.e. $w=32$ for \cipher{MORUS640} and $w=64$ for \cipher{MORUS1280}. The word-wise circular shift constants $b_i$ in round $i$ are defined in Table~\ref{Tbl:rcon}. Besides, the left rotation of a whole $q$-bit register is used. The register-wise circular shift constants are also listed in Table~\ref{Tbl:rcon}.
\begin{table}[!htb]
\centering
\caption{Rotation Constants of MORUS} \label{Tbl:rcon}
\begin{tabular}{c||ccccc|ccccc} \hline
& \multicolumn{5}{c|}{register-wise circular shift} & \multicolumn{5}{c}{word-wise circular shift} \\
                  & \makebox[2em]{$b_0$} & \makebox[2em]{$b_1$} & \makebox[2em]{$b_2$} & \makebox[2em]{$b_3$} & \makebox[2em]{$b_4$} & \makebox[2em]{$b'_0$} & \makebox[2em]{$b'_1$} & \makebox[2em]{$b'_2$} & \makebox[2em]{$b'_3$} & \makebox[2em]{$b'_4$} \\ \hline
\cipher{MORUS640} &  5 & 31 &  7 & 22 & 13 & 32 &  64 &  96 &  64 & 32 \\
\cipher{MORUS1280}& 13 & 46 & 38 &  7 &  4 & 64 & 128 & 192 & 128 & 64 \\ \hline
\end{tabular}
\end{table}
We denote those register-wise and word-wise left circular shift by $x \rotl y$ and $x \rotlxy y$, respectively. (Those are denoted by \texttt{<<<} and \texttt{Rotl\_xxx\_yy} respectively in the \cipher{MORUS} design document.)

Let $\cdot$ denote the bit-wise AND operation. $S^{t+1} \leftarrow {\tt StateUpdate}(S^t,m^t)$ is defined as follows.
\begin{align*}
\textrm{Round 1:} &&
S^{t+1}_0 &\leftarrow ( S_t^0 \oplus S^t_1 \cdot S^t_2 \oplus S^t_3 ) \lll_w b_0 &
S^t_3 \leftarrow S^t_3 \lll b'_0.\\
\textrm{Round 2:} &&
S^{t+1}_1 &\leftarrow ( S_t^1 \oplus S^t_2 \cdot S^t_3 \oplus S^t_4 \oplus m_i ) \lll_w b_1 &
S^t_3 \leftarrow S^t_3 \lll b'_1.\\
\textrm{Round 3:} &&
S^{t+1}_2 &\leftarrow ( S_t^2 \oplus S^t_3 \cdot S^t_4 \oplus S^t_0 \oplus m_i ) \lll_w b_2 &
S^t_3 \leftarrow S^t_3 \lll b'_2.\\
\textrm{Round 4:} &&
S^{t+1}_3 &\leftarrow ( S_t^3 \oplus S^t_4 \cdot S^t_0 \oplus S^t_1 \oplus m_i ) \lll_w b_3 &
S^t_3 \leftarrow S^t_3 \lll b'_3.\\
\textrm{Round 5:} &&
S^{t+1}_4 &\leftarrow ( S_t^4 \oplus S^t_0 \cdot S^t_1 \oplus S^t_2 \oplus m_i ) \lll_w b_4 &
S^t_3 \leftarrow S^t_3 \lll b'_4.
\end{align*}

\subsubsection{Initialization.}
The initialization of \cipher{MORUS640} consists of loading a 128-bit key $K$ and a 128-bit nonce IV into the state, and running the {\tt StateUpdate} for 16 steps. The key and IV are loaded into the state as follows:
\begin{align*}
S^{-16}_0 = \textrm{IV}_{128}, &&
S^{-16}_1 = K_{128}, &&
S^{-16}_2 = 1^{128}, &&
S^{-16}_3 = const_0, &&
S^{-16}_4 = const_1.
\end{align*}
Then the internal state is updated 16 times by  ${\tt StateUpdate}(S^t,0)$ for $t=-16,-15,\cdots,-1$. Finally, the key is xored to the state again as $S^0_1 \leftarrow S^0_1 \oplus K$.

The initialization of \cipher{MORUS1280} is slightly different from that of \cipher{MORUS640} due to the different register size. The state is initialized as
\begin{align*}
S^{-16}_0 = \textrm{IV}_{128} \| 0^{128}, &&
S^{-16}_1 = K_{128}\|K_{128}, &&
S^{-16}_2 = 1^{256}, \\
S^{-16}_3 = 0^{256}, &&
S^{-16}_4 = const_1\|const_0.
\end{align*}
After updating the state 16 times as \cipher{MORUS640}, the state is updated by $S^0_1 \leftarrow S^0_1 \oplus K_{128}\|K_{128}$.

\subsubsection{Associated Data Processing.}
After the initialization, the associated data $A$ is processed. For the padding, if the last associated data block is not a full block, use `0' bits to pad it to 128 bits and 256 bits for \cipher{MORUS640} and \cipher{MORUS1280}, respectively, and the padded full block is used to update the state. Note that if the length of $A$ dented by `$adlen$' is 0, the associated data processing is skipped.

Then, state is updated by $S^{t+1} \leftarrow {\tt StateUpdate}(S^t, A^t)$ for $t=0,1,\cdots,\ell$ where $\ell = \lceil \frac{adlen}{128} \rceil -1$ and $\ell = \lceil \frac{adlen}{256} \rceil -1$ for \cipher{MORUS640} and \cipher{MORUS1280}, respectively.

\subsubsection{Encryption.}
At each step of the encryption, a 16-byte (resp. 32-byte) message block $M_i$ is used to update the state, and $M_t$ is encrypted to $C_t$ in \cipher{MORUS640} (resp. \cipher{MORUS1280}).

If the last message block is not a full block, use `0' bits to pad it to 128 and 256 bits for \cipher{MORUS640} and \cipher{MORUS1280}, respectively, and the padded full block is used to update the state. But only the partial block is encrypted. Note that if the message length denoted by `$msglen$' is 0, encryption is skipped.

In \cipher{MORUS640}, Let $u = \lceil \frac{adlen}{128} \rceil$ and $v = \lceil \frac{msglen}{128} \rceil$. The following is performed for $t=0, 1, \cdots, v-1$;
\begin{align*}
C^t &\leftarrow M^t \oplus S^{u+t}_0 \oplus (S^{u+t}_1 \lll 96) \oplus (S^{u+t}_2 \cdot S^{u+t}_3),\\
S^{u+t+1} &\leftarrow {\tt StateUpdate}(S^{u+t},M^t).
\end{align*}

For \cipher{MORUS1280}, $u = \lceil \frac{adlen}{256} \rceil$ and $v = \lceil \frac{msglen}{256} \rceil$. The equation to compute ciphertext block $C^t$ is $C^t \leftarrow M^t \oplus S^{u+t}_0 \oplus (S^{u+t}_1 \lll 192) \oplus (S^{u+t}_2 \cdot S^{u+t}_3)$.

\subsubsection{Finalization.}
The finalization generates the authentication tag $T$ using 10 more steps. We only discuss the case that $T$ is not truncated. The associated data length and the message length are used to update the state as follows.
\begin{enumerate}
\item $tmp \leftarrow adlen\|msglen$ for \cipher{MORUS640} and $tmp \leftarrow adlen\|msglen\|0^{128}$ for \cipher{MORUS1280}, where $adlen$ and $msglen$ are represented as 64-bit integers.
\item $S^{u+v}_4 \leftarrow S^{u+v}_4 \oplus S^{u+v}_0.$
\item For $t = u+v, u+v+1, \cdots, u+v+9,$ compute
%\begin{equation*}
$S^{t+1} \leftarrow {\tt StateUpdate} (S^t, tmp)$.
%\end{equation*}
\item $T = S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 96) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$ for \cipher{MORUS640} and $T = S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 192) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$ for \cipher{MORUS1280}.
\end{enumerate}

\subsubsection{Security.}
When the tag size is 128 bits, integrity is claimed up to 128-bits and confidentiality is claimed up to the number of bits of the key.
\begin{table}[!htb]
\centering
\caption{Security Goals of MORUS.}
\begin{tabular}{ccc}\hline
                        & Confidentiality (bits) & Integrity (bits) \\ \hline
\cipher{MORUS640}-128   & 128                    & 128              \\
\cipher{MORUS1280}-128  & 128                    & 128              \\
\cipher{MORUS1280}-256  & 256                    & 128              \\ \hline
\end{tabular}
\end{table}

\subsection{Notation}

@@

%Finally, let $x$ denote a register. Recall that registers of \cipher{MORUS} are composed of four words. We now define register-wise and word-wise rotations (denoted by \texttt{<<<} and \texttt{Rotl\_xxx\_yy} respectively in the \cipher{MORUS} design document):\\
%\centerline{
%\def\arraystretch{1.25}
%\begin{tabular}{lcl}
%$x \rotl y$ &:& circular shift of $x$ by $y$ bits to the left.\\
%$x \rotlxy y$ &:& circular shift of each word within $x$ by $y$ bits to the left.\\
%\end{tabular}
%}
%
%The $b_i$'s denote rotation constants, whose value depends on which version of \cipher{MORUS} is considered. Throughout, $w$ denotes the  word size, i.e. $w=32$ for \cipher{MORUS640} and $w=64$ for \cipher{MORUS1280}.

@@

\section{Analysis on Initialization and Finalization of Reduced MORUS}

\subsection{Forgery with Reduced Finalization}
\subsubsection{Overview.}
\subsubsection{Differential Trails.}

\subsection{Extending State Recovery to Key Recovery}
\subsubsection{Overview.}
\subsubsection{Full Diffusion Rounds.}
